The Precambrian paleogeography of Laurentia

### 2021-06-01 Manuscript/Laurentia_Paleogeo_Manuscript.tex
small update to Laurentia rifted margin text

On the southern part of the eastern margin, a record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig. \ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a, Whitmeyer2007a}. The ages of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}. In this model, spatially-restricted rifting occurs ca. 760 to 680 Ma in the region of present-day North Carolina and Virginia. Rifting initiated in the region from present-day New York to Newfoundland ca. 620 to 580 Ma and by ca. 580 to 550 Ma rifting extended along the length of Laurentia's eastern margin. The last phase of this rifting has been interpreted to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane ca. 550 Ma \citep{Dickerson1998a, Martin2019a}. As with other rifts, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting and departure of a major craton, as the record that lingers on the craton is similar. Recognizing this ambiguity, it has been proposed that rather than being associated with spatially-restricted or failed rifting, ca. 700 Ma extension was associated with breakup and separation of a conjugate continent and that there was subsequent rifting of smaller terranes off the eastern Laurentia margin ca. 600 Ma \citep{Chew2008a, Escayola2011a, Robert2020a}. In these models, the separation of a conjugate continent ca. 700 Ma (interpreted to be Amazonia) leads to the formation of an ocean termed the Puncoviscana Ocean by \citet{Escayola2011a} and the Paleo-Iapetus Ocean by \citet{Robert2020a}. The subsequent rifting of terranes off the Laurentia margin ca. 600 Ma would have resulted in the formation of the (Neo-)Iapetus Ocean and the record of thermal subsidence associated with passive margin development on Laurentia \citep{Escayola2011a, Robert2020a}.

>>>>

On the southern part of the eastern margin, a record of this rifting is preserved as rift basins that were part of failed arms (Rome trough, Reelfoot rift and Oklahoma aulacogen; Fig. \ref{fig:Laurentia_map}) as well as prolonged Cambrian to Ordovician passive margin thermal subsidence along the margin \citep{Bond1984a, Whitmeyer2007a}. The ages of igneous intrusions that have been interpreted to be rift-related play a significant role in interpretations of this history such as in the rift development model of \citet{Burton2010a}. In this model, spatially-restricted rifting occurs ca. 760 to 680 Ma in the region of present-day North Carolina and Virginia. Rifting initiated in the region from present-day New York to Newfoundland ca. 620 to 580 Ma and by ca. 580 to 550 Ma rifting extended along the length of Laurentia's eastern margin. The last phase of this rifting has been interpreted to be associated with the separation of the Argentine pre-Cordillera Cuyania terrane ca. 550 Ma \citep{Dickerson1998a, Martin2019a}. Rifting of the Laurentia margin north of the New York promontory was occuring by ca. 560 Ma based on the age of volcanics that are interlayered with sedimentary rocks interpreted to be synrift deposits \citep{}Hodych2007.

As with other rifted margins, it is difficult to distinguish the separation of a cratonic fragment as a microcontinent from the rifting and departure of a major craton, as the record that lingers on the craton is similar. Recognizing this ambiguity, it has been proposed that rather than being associated with spatially-restricted or failed rifting, ca. 700 Ma extension was associated with breakup and separation of a conjugate continent and that there was subsequent rifting of smaller terranes off the eastern Laurentia margin ca. 600 Ma \citep{Chew2008a, Escayola2011a, Robert2020a}. In these models, the separation of a conjugate continent ca. 700 Ma (interpreted to be Amazonia) leads to the formation of an ocean termed the Puncoviscana Ocean by \citet{Escayola2011a} and the Paleo-Iapetus Ocean by \citet{Robert2020a}. The subsequent rifting of terranes off the Laurentia margin ca. 600 Ma would have resulted in the formation of the (Neo-)Iapetus Ocean and the record of thermal subsidence associated with passive margin development on Laurentia \citep{Escayola2011a, Robert2020a}.
